
We provide here a comprehensive inventory of snoRNAs in fungi together with
a detailed analysis of the evolution of snoRNA families and their target
specificities. The investigation of 147 different taxa provides a detailed
history of gain, loss, and duplication events for 68 families of box C/D
snoRNAs and 50 families of box H/ACA snoRNAs involving more than 7,700
individual snoRNA sequences. \PFS{For 18 snoRNA families previously
unrecognized homology with other families has been uncovered. These data
constitute a substantial extension and refinement of the accumulated
knowledge on \sno{}s. Data and refined models will become available in the
\rfam\ database and collectively form an important step towards a global
understanding of the evolution of the snoRNAome. Since our approach is
based on homology search, it is fundamentally limited by the seed sequences
that have been observed and classified as \sno{}s in at least one
organism. It is very unlikely, therefore, that this study presents a
complete picture despite doubling the number of \sno\ families and
increasing the number \sno\ sequences by more than a factor of four.}


The processing of this amount data is well beyond the realm of manual
curation and has been possible only with the help of \snostrip, a pipeline
specifically developed to investigate the evolution of snoRNA families
across a broad phylogenetic range \cite{Bartschat:2014}.  The in-depth
analysis of potential target interactions adds a new layer of
information. We have demonstrated here that the coevolution of \sno s and
their targets can be tracked with high resolution based on the functional
characteristics of the \sno s as determined by \snostrip\ together with a
quantitative assessment of predicted RNA-RNA interaction based on the the
Interaction Conservation Index (ICI) \cite{Kehr:2014}.

Similar to Metazoa, fungal \haca s show a higher loss-ratio compared
to \cd s. This might have both a technical and a biological
explanation that manifests itself on two different levels. Since \haca
s do not share long ASEs but rather short bipartite pseudouridylation
pockets, it becomes considerably harder to detect homologous \sno s
over large evolutionary timescales. This effect may limit the
scope of the homology search procedure. The short interacting regions
also make these molecules more vulnerable to mutations that disrupting
the snoRNA-targetRNA interaction. At the same time, the presence of
the second, independent ASE in the other hairpin may be a sufficient
cause to retain mutated genes.

In general, fungal \sno s have well preserved target interactions and
most families are found to contain exactly one highly conserved
anti-sense element. The remaining target region is in turn free to
evolve or to adapt to new lineage-specific or even species-specific
targets. Here we introduced a variation on the ICI scoring adapted to
subclades, allowing a much more detailed quantitative assessment of
target turnover. Many of the predictions made here of course await
experimental validation, given that experimental evidence for
RNA-target interactions as well as direct measurements of chemical
modifications in the primary target molecules (rRNAs and snRNAs) are
still restricted to a few model organisms.

The computational analysis reported here strongly suggests that
snoRNAs not only address a highly conserved ASE but also frequently
have additional, secondary targets. The possibility that a single
\sno\ target site exerts two distinct guiding functions has been
reported from budding yeast \haca s. The budding yeast \sno\ family
snR3 (HACA\_3), for example, is verified to target two modification
sites in its second hairpin \cite{Schattner:2004}.  Both interactions
can be tracked across the Dikarya.  Nevertheless, there is still very
little experimental data on the generality of this effect and most of
the predicted 'double' target sites will still require experimental
verification. Convincing examples of remarkably conserved multiple
interactions are found in \cd\ families snR40 (CD\_43) and snR70
(CD\_61), which exhibit two and five high-scoring target-interactions
at a single ASE, respectively.  These findings suggest the possibility
that \sno s are, at least under certain circumstances, able to guide
different modifications with the same ASE. This might be dependent on
developmental phases, or more complex mechanisms involving
conformational changes of the target.

In some cases of \haca s, these additional targets exhibit better ICI
scores than the annotated modification sites. Since the ICI combines
evidence from thermodynamic stability and evolutionary conservation, these
predictions cannot be easily dismissed as false positives. The specialized
ribosome hypothesis proposes distinct ribosomal conformations in different
developmental stages and stress levels that might also entail different
chemical modification patterns of the rRNAs; it is entirely plausible in
this scenario that some modifications and thus snoRNA interaction sites
have remained undetected \citep{Xue:2012}. The existence of
stress-induced conditional pseudouridylations indeed has been reported for
the for the U2 snRNA of budding yeast \cite{Wu:2011}. The snR81 RNA, which
is also responsible for guidance of a constitutive U2 pseudouridylation,
guides one of the novel modifications through imperfect and redundant base
pairing. The authors speculate that conditionally induced modifications in
RNA may well be a rather frequent phenomenon.

On the other hand, we also found convincing evidence that some
modifications are guided by two, three, or even more \sno\
families. First, this includes redundant guides, meaning that two
snoRNA families of the same species are responsible for the same
modification. Second, we observed several target sites that are
addressed by different snoRNA families in different taxonomic
groups. A good example for the latter situation is the predicted
pseudouridine at position 5.8S-18. Although there is not direct
experimental evidence that this particular position is modified
\emph{in vivo}, the site is predicted as a target for several distinct
snoRNA families by \snoop\ (see supplement section S21).

The fact that specific modification sites are predicted to be guided
by more than just one \sno\ family in the same organism has several
possible reasons. SnoRNA expression recently was reported to be
strongly regulated in development and between tissues or cell lines
\cite{Kapushesky:2012, Jorjani:2016}. It may thus be necessary for the
organism to compensate for snoRNAs that are lowly expressed under
certain circumstances to maintain the functional modification levels
of the target RNA. This may be achieved by means of paralog or through
the redundant target binding capability of another \sno\ family.

In summary we observe that the landscape of snoRNAs keeps changing
over the the evolutionary time-scale of the kingdom Fungi. We observe
both the extinction of entire \sno\ families and the innovation of new
ones. The function of snoRNA families itself also changes at these
evolutionary scales, showing loss, gain, and turn-over of guiding
functions that lead to target switches. The number of known snoRNA
families in Fungi is lower than in animals, correlating well with the
observation that animals have more (reported) modification sites in
their rRNAs and snRNAs than ``lower'' Eukaryotes (see
\texttt{Modomics} and the \texttt{RNA Modification Database}
\cite{Machnicka:2013, Cantara:2011}) or even Bacteria (which have
target-specific enzymes for each individual modification instead of
the generic enzyme machinery with snoRNAs as evolutionary flexible
``address labels'').

On the other hand, there are many similarities between the fungal and
the metazoan snoRNAome.  A common feature is the detectable burst in
the \sno\ diversity at each major branching point in the taxonomic
tree of both kingdoms.  In case of \cd s, the distribution of orphan,
single guided, and double guided \snos\ is quite similar compared
between fungi and animals, as reported by the human \sno\ atlas
\citep{Jorjani:2016}: Over 70\% of the human box C/D carrying \sno s
are found to be single guided (75\% in Fungi).  In both human and
fungi the remainder is about equally split between double guided and
orphan snoRNAs. The situation is somewhat different for \haca s: in
human double guided \sno s comprise the largest group (47\%), while in
Fungi, only 22\% of the \haca\ families target two distinct
pseudouridylation sites with both hairpins.

