\PassOptionsToPackage{sort&compress}{natbib}
\documentclass[preprint,3p,times,twocolumn]{elsarticle}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{rotating,capt-of}
\usepackage[small]{caption}
\usepackage{booktabs}
\usepackage{fixltx2e} % two-column fig/tab float order
\usepackage{float} % for placing figures where i want
\usepackage{afterpage}
\usepackage{soul}  % KR: added as it contains \st command
\usepackage{epsfig, a4wide}
\usepackage{floatflt}

\newcommand{\myparagraph}[1]{
  \paragraph*{\normalfont\itshape #1}\hspace{5pt}}

% strange snos
\definecolor{purple}{RGB}{180,90,200}
\definecolor{dgreen}{RGB}{0,160,0}
\definecolor{turquoise}{RGB}{0,180,140}
\renewcommand\dblfloatpagefraction{0.03}
\renewcommand\topfraction{.95}
\renewcommand\bottomfraction{.95}
\renewcommand\textfraction{.05}
\renewcommand\floatpagefraction{.95}
\renewcommand\dbltopfraction{.95}
\renewcommand\dblfloatpagefraction{.95}
\newcommand{\PFS}[1]{\begingroup\color{blue}#1\endgroup}
%\newcommand{\DOM}[1]{\begingroup\color{green}#1\endgroup}
\newcommand{\JH}[1]{\begingroup\color{purple}#1\endgroup}
%\newcommand{\KR}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\SB}[1]{\begingroup\color{turquoise}#1\endgroup}
\newcommand{\TODO}[1] {\begingroup\color{red}#1\endgroup}
%\newcommand{\FINAL}[1]{\begingroup\color{dgreen}#1\endgroup}
\newcommand{\ACC}[1]{\emph{\textbf{#1}}}
\newcommand{\s}[1]{\begin{tiny}#1\end{tiny}}
\newcommand{\url}[1]{\texttt{\small #1}}
\newcommand{\maxentscan}{\texttt{MaxEntScan}}

% Sebastian (Diss)
%% abbreviations
\newcommand{\snos}{snoRNAs}
\newcommand{\sno}{snoRNA}
\newcommand{\cd}{box C/D snoRNA}
\newcommand{\haca}{box H/ACA snoRNA}
%% programs
\newcommand{\blast}{\texttt{Blast}}
\newcommand{\snostrip}{\texttt{snoStrip}}
\newcommand{\epope}{\texttt{ePoPE}}
\newcommand{\plexy}{\texttt{PLEXY}}
\newcommand{\snoop}{\texttt{RNAsnoop}}
\newcommand{\fold}{\texttt{RNAfold}}
\newcommand{\alifold}{\texttt{RNAalifold}}
\newcommand{\locarna}{\texttt{LocaRNA}}
%% databases
\newcommand{\ncbi}{\texttt{NCBI}}
\newcommand{\umass}{\texttt{UMass}}



%FUNGI
\newcommand{\afu}{\emph{A.fumigatus}}
\newcommand{\Afu}{\emph{Aspergillus fumigatus}}
\newcommand{\nfi}{\emph{N.fischeri}}
\newcommand{\Nfi}{\emph{Neosartorya fischeri}}
\newcommand{\calb}{\emph{C.albicans}}
\newcommand{\Calb}{\emph{Candida albicans}}
\newcommand{\schizo}{\emph{Schizosaccharomyces}}
\newcommand{\fusarium}{\emph{Fusarium}}
\newcommand{\neurospora}{\emph{Neurospora}}
\newcommand{\trichoderma}{\emph{Trichoderma}}
\newcommand{\botrytis}{\emph{Botrytis}}
\newcommand{\sclerotinia}{\emph{Sclerotinia}}
\newcommand{\aspergillus}{\emph{Aspergillus}}
\newcommand{\penicillium}{\emph{Penicillium}}
\newcommand{\saccharomyces}{\emph{Saccharomyces}}
\newcommand{\dothi}{Dothideomycetes}
\newcommand{\tde}{\emph{T.deformans}}
\newcommand{\Tde}{\emph{Taphrina deformans}}
\newcommand{\Spo}{\emph{Schizosaccharomyces pombe}}
\newcommand{\spo}{\emph{S.pombe}}
\newcommand{\Sja}{\emph{Schizosaccharomyces japonicus}}
\newcommand{\sja}{\emph{S.japonicus}}
\newcommand{\scm}{\emph{S.complicata}}
\newcommand{\Scm}{\emph{Saitoella complicata}}
\newcommand{\Ssl}{\emph{Sclerotinia sclerotiorum}}
\newcommand{\ssl}{\emph{S.sclerotiorum}}
\newcommand{\Clo}{\emph{Chalara longipes}}
\newcommand{\clo}{\emph{C.longipes}}
\newcommand{\Pan}{\emph{Podospora anserina}}
\newcommand{\pan}{\emph{P.anserina}}
\newcommand{\Oma}{\emph{Oidoidendron maius}}
\newcommand{\oma}{\emph{O.maius}}
\newcommand{\Fgr}{\emph{Fusarium graminearum}}
\newcommand{\fgr}{\emph{F.graminearum}}
\newcommand{\fve}{\emph{F.verticillioides}}
\newcommand{\Fve}{\emph{Fusarium verticillioides}}
\newcommand{\Fox}{\emph{Fusarium oxysporum}}
\newcommand{\fox}{\emph{F.oxysporum}}
\newcommand{\alternaria}{\emph{Alternaria}}
\newcommand{\Abr}{\emph{Alternaria brassicicola}}
\newcommand{\abr}{\emph{A.brassicicola}}
\newcommand{\mycosphaerella}{\emph{Mycosphaerella}}
\newcommand{\Mpi}{\emph{Mycosphaerella pini}}
\newcommand{\mpi}{\emph{M.pini}}
\newcommand{\pex}{\emph{Penicillium expansum}}
\newcommand{\pcy}{\emph{Penicillium chrysogenum}}
\newcommand{\Rha}{\emph{Rhodotorula hasegawae}}
\newcommand{\rha}{\emph{R.hasegawae}}
\newcommand{\Rda}{\emph{Rhodosporidium dacryoidum}}
\newcommand{\rda}{\emph{R.dacryoidum}}
\newcommand{\Rgr}{\emph{Rhodotorula graminis}}
\newcommand{\rgr}{\emph{R.graminis}}
\newcommand{\Rmi}{\emph{Rhodotorula minuta}}
\newcommand{\rmi}{\emph{R.minuta}}
\newcommand{\Sli}{\emph{Sporobolomyces linderae}}
\newcommand{\sli}{\emph{S.linderae}}
\newcommand{\Cne}{\emph{Cryptococcus neoformans}}
\newcommand{\cne}{\emph{C.neoformans}}
\newcommand{\Ncr}{\emph{Neurospora crassa}}
\newcommand{\ncr}{\emph{N.crassa}}
\newcommand{\Nte}{\emph{Neurospora tetrasperma}}
\newcommand{\nte}{\emph{N.tetrasperma}}
\newcommand{\Andi}{\emph{Aspergillus nidulans}}
\newcommand{\andi}{\emph{A.nidulans}}
\newcommand{\Sce}{\emph{Saccharomyces cerevisiae}}
\newcommand{\sce}{\emph{S.cerevisiae}}
\newcommand{\Spa}{\emph{Saccharomyces paradoxus}}
\newcommand{\spa}{\emph{S.paradoxus}}
\newcommand{\Smi}{\emph{Saccharomyces mikatae}}
\newcommand{\smi}{\emph{S.mikatae}}
\newcommand{\Kpa}{\emph{Komagataella pastoris}}
\newcommand{\kpa}{\emph{K.pastoris}}
\newcommand{\Cpr}{\emph{Cryphonectria parasitica}}
\newcommand{\cpr}{\emph{C.parasitica}}
\newcommand{\Glo}{\emph{Glarea lozoyensis}}
\newcommand{\glo}{\emph{G.lozoyensis}}
\newcommand{\Yli}{\emph{Yarrowia lipolytica}}
\newcommand{\yli}{\emph{Y.lipolytica}}
\newcommand{\Ecu}{\emph{Encephalitozoon cuniculi}}
\newcommand{\ecu}{\emph{E.cuniculi}}
\newcommand{\Ror}{\emph{Rhizopus oryzae}}
\newcommand{\ror}{\emph{R.oryzae}}
\newcommand{\Tre}{\emph{Trichoderma reesei}}
\newcommand{\tre}{\emph{T.reesei}}
\newcommand{\Npa}{\emph{Nematocida parisii}}
\newcommand{\npa}{\emph{N.parisii}}
\newcommand{\Eae}{\emph{Edhazardia aedis}}
\newcommand{\eae}{\emph{E.aedis}}
\newcommand{\Vco}{\emph{Vittaforma corneae}}
\newcommand{\vco}{\emph{V.corneae}}
\newcommand{\Mbi}{\emph{Metschnikowia bicuspidata}}
\newcommand{\mbi}{\emph{M.bicuspidata}}
\newcommand{\Mva}{\emph{Meliniomyces variabilis}}
\newcommand{\mva}{\emph{M.variabilis}}
\newcommand{\Bci}{\emph{Botrytis cinerea}}
\newcommand{\bci}{\emph{B.cinerea}}
\newcommand{\Val}{\emph{Verticillium alfalfae}}
\newcommand{\val}{\emph{V.alfalfae}}
\newcommand{\Vda}{\emph{Verticillium dahliae}}
\newcommand{\vda}{\emph{V.dahliae}}
\newcommand{\Sal}{\emph{Sodiomyces alkalinus}}
\newcommand{\sal}{\emph{S.alkalinus}}
\newcommand{\Aac}{\emph{Acremonium alcalophilum}}
\newcommand{\aac}{\emph{A.~alcalophilum}}
\newcommand{\Nha}{\emph{Nectria haematococca}}
\newcommand{\nha}{\emph{N.haematococca}}
\newcommand{\Mth}{\emph{Myceliophthora thermophila}}
\newcommand{\mth}{\emph{M.thermophila}}
\newcommand{\Cgl}{\emph{Chaetomium globosum}}
\newcommand{\cgl}{\emph{C.globosum}}
\newcommand{\Tar}{\emph{Thielavia arenaria}}
\newcommand{\tar}{\emph{T.arenaria}}
\newcommand{\Tte}{\emph{Thielavia terrestris}}
\newcommand{\tte}{\emph{T.terrestris}}
\newcommand{\Ztr}{\emph{Zymoseptoria tritici}}
\newcommand{\ztr}{\emph{Z.tritici}}
\newcommand{\Tvi}{\emph{Trichoderma virens}}
\newcommand{\tvi}{\emph{T.virens}}
\newcommand{\Bco}{\emph{Baudoinia compniacensis}}
\newcommand{\bco}{\emph{B.compniacensis}}
\newcommand{\Kla}{\emph{Kluyveromyces lactis}}
\newcommand{\kla}{\emph{K.lactis}}
\newcommand{\Cga}{\emph{Candida glabrata}}
\newcommand{\cga}{\emph{C.glabrata}}
\newcommand{\Pme}{\emph{Pichia membranifaciens}}
\newcommand{\pme}{\emph{P.membranifaciens}}
\newcommand{\Ppl}{\emph{Postia placenta}}
\newcommand{\ppl}{\emph{P.placenta}}
\newcommand{\Nfu}{\emph{Nadsonia fulvescens}}
\newcommand{\nfu}{\emph{N.fulvescens}}
\newcommand{\Asp}{\emph{Atractiellales sp}}
\newcommand{\asp}{\emph{A.sp}}
\newcommand{\Bma}{\emph{Bipolaris maydis}}
\newcommand{\bma}{\emph{B.maydis}}
\newcommand{\Ptt}{\emph{Pyrenophora tritici-repentis}}
\newcommand{\ptt}{\emph{P.tritici-repentis}}
\newcommand{\Ana}{\emph{Aureobasidium namibiae}}
\newcommand{\ana}{\emph{A.namibiae}}
\newcommand{\Cbb}{\emph{Cucurberita berberidis}}
\newcommand{\Pno}{\emph{Parastagonospora nodorum}}
\newcommand{\wse}{\emph{W.sebi}}
\newcommand{\Wse}{\emph{Wallemia sebi}}
\newcommand{\opi}{\emph{O.piceae}}
\newcommand{\Opi}{\emph{Ophiostoma piceae}}
\newcommand{\cdu}{\emph{C.dubliniensis}}
\newcommand{\Cdu}{\emph{Candida dubliniensis}}
\newcommand{\ctr}{\emph{C.tropicalis}}
\newcommand{\Ctr}{\emph{Candida tropicalis}}


%Trennung verhindern
\hyphenation{snoRNAs}
\hyphenation{snoRNA}
\hyphenation{microRNA}
\hyphenation{microRNAs}
\hyphenation{miRNA}
\hyphenation{miRNAs}
\hyphenation{lncRNA}
\hyphenation{lncRNAs}
\hyphenation{lincRNA}
\hyphenation{lincRNAs}
\hyphenation{di-nucleo-tide}


\journal{Preprint}

\DeclareCaptionLabelFormat{simplesupp}{#1~S#2} % new caption format with Sxx

\begin{document}

\begin{frontmatter}

\title{The Fungal snoRNAome}

\author[LEI]{Sebastian Canzler\corref{cor1}}
\ead{sebastian@bioinf.uni-leipzig.de}
\author[LEI,IZBI,MPI,IZI,RTH,TBI,SFI]{Peter F.\ Stadler}
\ead{studla@bioinf.uni-leipzig.de}
\author[UFZ]{Jana Hertel}
\ead{jana.hertel@ufz.de}  

\address[LEI]{Bioinformatics Group, Department of Computer Science,
  Leipzig University,
  H{\"a}rtelstra{\ss}e 16-18, D-04107 Leipzig, Germany
}
\address[IZBI]{Interdisciplinary Center for Bioinformatics,
  German Centre for Integrative Biodiversity Research (iDiv)
  Halle-Jena-Leipzig, Competence Center for Scalable Data Services
  and Solutions, and Leipzig Research Center for Civilization Diseases,
  Leipzig University
}
\address[UFZ]{Young Investigators Group Bioinformatics and Transcriptomics, 
  Department Proteomics, 
  Helmholtz Centre for Environmental Research -- UFZ, 
  Permoserstra{\ss}e 15,
  D-04318 Leipzig, Germany
}
\address[IZI]{Department of Diagnostics, 
  Fraunhofer Institute for Cell Therapy and Immunology -- IZI,
  Perlickstra{\ss}e 1, 
  D-04103 Leipzig, Germany
}
\address[MPI]{Max Planck Institute for Mathematics in the Sciences,
  Inselstra{\ss}e 22, D-04103 Leipzig, Germany
}
\address[TBI]{Department of Theoretical Chemistry,
  University of Vienna,
  W{\"a}hringerstra{\ss}e 17, A-1090 Wien, Austria
}
\address[RTH]{Center for non-coding RNA in Technology and Health,
  University of Copenhagen, Gr{\o}nneg{\aa}rdsvej 3, 
  DK-1870 Frederiksberg C, Denmark
}
\address[SFI]{Santa Fe Institute, 1399 Hyde Park Rd., Santa Fe, NM 87501,
  USA\\[1em]
}

%\cortext[jfa]{Joint first authors}
\cortext[cor1]{Corresponding author}

% --------------------------------------------------------------------------- %


\begin{abstract}
  Small nucleolar RNAs (snoRNAs) are essential players in the rRNA
  biogenesis due to their involvement in the nucleolytic processing of the
  precursor and the subsequent guidance of nucleoside modifications. Within
  the kingdom Fungi, several species-specific surveys have explored their
  snoRNA repertoire. However, the wide range of the snoRNA landscape
  spanning all major fungal lineages has not been mapped so far, mainly
  because of missing tools for automatized \sno\ detection and functional
  analysis. Here, we report a comprehensive inventory of fungal {\sno}s
  together with an in-depth investigation of their evolutionary history
  including innovations, deletions, and target switches. This large-scale
  analysis, incorporating more than 120 \sno\ families with more than 7700
  individual \sno\ sequences, shows that the {\sno} landscape is subject
  to ongoing re-arrangements and adaptations, e.g., through
  lineage-specific targets and redundant guiding functions.

  An electronic supplement containing the data sets used and produced in
  this study is available at
  \url{http://www.bioinf.uni-leipzig.de/publications/supplements/17-001}. 
\end{abstract}

\begin{keyword}
  small nucleolar RNAs, snoRNA, fungi, evolution, target switch, snoRNA target, 	
  conservation	
\end{keyword}

\end{frontmatter}

% --------------------------------------------------------------------------- %

\section{Introduction}

Small nucleolar RNAs (snoRNAs) are non-protein-coding RNAs (ncRNAs) that
guide the chemical modification of single nucleotides in other RNA
molecules.  Localized in the nucleolus of eukaryotic (and some archaeal)
cells, they associate with at least four proteins to form the small
nucleolar ribonucleoprotein (snoRNP) complex \cite{Reichow:2007}.  The
target RNA molecule is held in the correct position by base pairing to
short unpaired region(s) within the snoRNA usually referred to as the
antisense elements (ASE). The base pairing completely specifies the target
nucleotide. Known modifications are mostly located in ribosomal RNAs
(rRNAs) and small nuclear RNAs (snRNAs)
\cite{Decatur:2002,Darzacq:2002,Bratkovi:2011}, although some snoRNAs have
been shown to target residues in other RNA molecules such as transfer RNAs
\cite{Clouet_d'Orval:2001,Dennis:2001}, spliced leader RNAs
\cite{Uliel:2004}, or brain-specific messenger RNAs \cite{Cavaillé:2000,
  Kishore:2006}. Furthermore, {\sno}s are known to be involved in the
nucleolytic processing of rRNA precursors, the synthesis of telomeric DNA,
genomic imprinting, and alternative splicing
\cite{Maxwell:1995,Tollervey:1997,Kiss:2002,Matera:2007}.

There are two distinct classes of snoRNAs: box C/D and box H/ACA
snoRNAs.  They are distinguished by their secondary structure,
sequence features, and the modification that they guide
\cite{Balakin:1996,Tollervey:1997}.  Box C/D snoRNAs form a stem-loop
structure with a rather long loop which is stabilized by the
associated proteins and guide the 2'-O-methylation of ribose groups.
Box H/ACA snoRNAs are longer, fold into a thermodynamically more
stable double stem-loop structure, and guide pseudouridylation of
uracil residues in the target RNA.  In addition to these two classes,
there are chimeric snoRNAs that share features of both classes, are
much longer and/or are described to have different functions
\cite{Darzacq:2002}. Similar to other small ncRNAs, snoRNAs require
both specific secondary structures and characteristic sequence motifs
to perform their function. These features are therefore preserved during
evolution and are clearly recognizable by comparative methods
\cite{Ganot:1997,Tollervey:1997}. While sequence motifs involved in
protein binding are common to all members of each of the the two
classes, the ASEs are conserved only among members of snoRNA families
with the same target. Overall, therefore, snoRNA sequences evolve
rapidly, making them hard to identify by purely sequence-base methods
such as \texttt{blast} \cite{Altschul:1990}.

To overcome this limitation, we introduced a computational annotation
pipeline \snostrip\ \cite{Bartschat:2014} that is specifically designed to
track all specific characteristics of snoRNAs. Here we use this approach to
analyze a large set of fungal species with genomes that are available in
decent quality for their snoRNA abundance.  We started with experimentally
verified snoRNAs in five fungi.  Further, we studied the evolutionary
conservation of those snoRNAs and the coevolution of snoRNAs and their
targets. We provide a comprehensive set of fungal snoRNAs, their detailed
description with respect to genomic location, box motifs and position,
potential/confirmed target information (including observed target
switches), family assignment and a suggestion of the evolutionary history
of individual snoRNA families.  All data can be viewed in and downloaded
from our electronic supplement. Manually curated snoRNA family alignments
will be submitted to the \texttt{Rfam} database \cite{Nawrocki:2015}.

% --------------------------------------------------------------------------- %

\section{Materials and Methods}

\subsection{Genome and snoRNA Data} 

Genome sequences from 147 fungal species were downloaded from Ensembl
Genomes \cite{Kersey:2016}, JGI \cite{Nordberg:2014}, Broad Institute
(Fungal Genome Initiative) , and Candida Genome Database
\cite{Skrzypek:2017}.  An \ncbi-based taxonomic tree displaying the
relationship, genome source, and version for all fungal organisms in
this evolutionary survey is shown in the supplementary Figures S1.
For 63 out of the 147 species, most snoRNA sequences were already
retrieved in a previous study, mostly as test of \snostrip\
\cite{Bartschat:2014}.  In this earlier work, we started with
experimentally detected snoRNAs extracted from five surveys for \Ncr\
\cite{Liu:2009}, \Afu\ \cite{Joechl:2008}, \Calb\
\cite{Mitrovich:2010}, \Sce\ \cite{Piekna-Przybylska:2007}, and \Spo\
\cite{Li:2005}. An overview of the experimentally verified \snos\ and
the corresponding publications is compiled in Supplementary Table
S2. The nomenclature of snoRNAs is consistent across different
species. A dictionary relating the species-specific, traditional \sno\
names as they were used in the original publications and the internal
snoRNA family designations used by \snostrip\ is included as
Supplementary Table S3.  Here we use the results of
\cite{Bartschat:2014} as our starting point.  The initial set comprises
3564 snoRNA sequences assigned to 123 snoRNA families in the 63
species. It includes 231 experimentally validated \sno\ genes taken
from the five publications.

\subsection{Homology search}

The \snostrip\ pipeline \cite{Bartschat:2014} was applied to the set
of collected snoRNAs and the 147 fungal species in an iterative
manner. Starting with Pezizomycotina, followed by Saccharomycotina,
and other lineages towards the root of the phylogenetic tree. Each
time new (plausible) homologous snoRNAs were detected, the procedure
was repeated to decrease the number of false negatives until no novel
homologs were found anymore.

\subsection{Data curation}

The candidate snoRNAs identified by \snostrip\ were curated regarding the
automatically identified box motifs, plausible lengths, and the overall fit
of each \sno\ sequence in its respective family. To identify incorrectly
annotated box motifs, the conservation of all predicted boxes was checked
by a comparison of the start positions within the \sno\ family
alignment. Motifs that start at unconserved positions are most probably
false annotations and were re-adjusted to fit the \sno\ family-specific box
pattern and and box position.  Sequences, where re-adjusted C- or D-boxes
did not agree with the canonical box motif pattern, were removed from further
analysis.
Candidates sequences that are either too long or too short were mostly
  the consequence of misannotated box motifs because \snostrip\ cuts \sno\
genes based on their box motif positions. For these candidates, box motifs
were analyzed with respect to their conserved start positions.
Subsequently, sequences with re-adjusted box motifs were automatically
trimmed or enlarged.

\subsection{Box motifs, sequence and structure}

Box motifs were generated from all \snostrip-derived \sno\ candidates
and compared to canonical box motifs of yeast and vertebrate snoRNAs.
Sequence lengths and distances between all box motifs were collected
and compared. Secondary structure prediction was done using the \fold\
and \alifold\ programs from the Vienna RNA
Package\cite{Hofacker:1994}.

\subsection{Phylogenetic analysis}

To follow the evolution of the snoRNA families along the phylogenetic
tree we used the software \epope\cite{Hertel:2015}.  It implements a
variant of Sankoff’s parsimony algorithm using the Dollo variant that
excludes the loss and re-gain of a gene family along the same lineage
during evolution.  Innovation and deletion$/$loss$/$divergence events
are deduced and mapped to the branches of the phylogenetic tree.  The
\epope\ results are combined for \emph{all} \sno\ families using the
\texttt{ePoPE\_summarize.pl} tool that comes with the \epope\
distribution.

\subsection{Target prediction and analysis}

Target prediction is part of the \snostrip\ pipeline. There, the
computational tools \plexy\ and \snoop\ are employed to predict
targets for \cd s and \haca s, respectively \cite{Kehr:2011,
  Tafer:2010}.  SnoRNAs are investigated for single or double guide
potential based on these predictions and/or confirmed target
interactions. SnoRNAs that remain without target association are
considered orphan. SnoRNAs that are assigned to the same family but
show variance in their associated target are investigated manually for
a potential target switch.

\subsection{Lineage-specific conservation of target interactions}
To study the conservation of interactions, the targets for each
individual snoRNA sequence are initially predicted and subsequently
their conservation in other species is evaluated. To formally
investigate the conservation, the Interaction Conservation Index (ICI)
was developed by Kehr \emph{et al.} \cite{Kehr:2014}. In brief, the
conservation of the modification and the conservation in a specific
snoRNA family are calculated as follows:
\begin{equation}
  \begin{split}
    ICI_{mod}\:(t,s) & = \frac{1}{|O(s)|} \left( \sum_{k\in O(t,s)} \frac{\varepsilon(t,s,k)}{\bar\varepsilon(t,k)} \right) \\
    ICI_{sno}\:(t,s) & = \frac{1}{|O(s)|} \left( \sum_{k\in O(t,s)} \frac{\varepsilon(t,s,k)}{\hat\varepsilon(s,k)} \right)
  \end{split}
\end{equation}
Here, $ \varepsilon(t,s,k) = \min_{x\in X(t,s,k)}
E_{\textrm{mfe}}[x,y_{t,k}] $ is the most negative interaction minimum free
energy between a snoRNA $x$ of family $s$ and the target $t$ in species
$k$. The normalizations 
\begin{equation}
\begin{split}
\bar\varepsilon(t,k)=\sum_{s\in
  S(t,k)}\varepsilon(t,s,k)/|S(t,k)| \\
 \hat\varepsilon(s,k)=\sum_{t\in
  T(s,k)}\varepsilon(t,s,k)/|T(s,k)|
\end{split}
\end{equation}
are obtained by averaging over all predictions of target $t$ in
species $k$ or all targets $t$ of snoRNA $s$ in species $k$,
respectively.  Their normalized parameters are then summed up over all
species $k\in O(t,s)$ in which a prediction of target $t$ is found for
snoRNA family $s$. The sum is then normalized w.r.t.\ the number of
species $|O(s)|$ in which the snoRNA family $s$ is present. This
approach is particularly suitable for modification sites that are
present in a large set of analyzed organisms. In cases where a
potential target appears to be lineage-specific, the ICI score will
drop to rather low values due to the normalization score $1/O(s)$ that
represents all organisms sharing a homologous \sno\ of family $s$.

To appropriately investigate alternative or additional targets that
merely appear in a particular subset of organisms, the ICI score
calculation has to be adapted to take the particular phylogenetic
distribution of a target interaction into account. Therefore, the
normalization is restricted to the smallest phylogenetic or taxonomic
subtree that harbors all organisms that share prediction of target
$t$ in snoRNA family $s$.  Assume the overall taxonomic tree is
represented by a tree $T=(V,E)$ with root $\gamma$. The minimal
subtree $U_\tau = (V_\tau, E_\tau)$ with root $\tau$ shares the node
set
$V_\tau = \{\ v\ |\ \forall (v,u)\ ,\ u \in V_\tau\ :\ LCA_T(v,u) \in
V_\tau\ \}$ where $LCA_T(v,u)$ is the lowest common ancestor in tree
$T$ of both nodes $v$ and $u$. More precisely, the $LCA$ is the lowest
node, i.e., the farthest node from the root, that has both $v$ and $u$
as descendants.  Hence, the ICI scores in a particular subtree rooted
at $\tau$ can be calculated as follows:
\begin{equation}
  \begin{split}
    ICI_{mod,\tau}\:(t,s) & = \frac{1}{|O_\tau(s)|}  \left( \sum_{k\in O_\tau(t,s)} \frac{\varepsilon(t,s,k)}{\bar\varepsilon(t,k)} \right)
    \\
    ICI_{sno,\tau}\:(t,s) & = \frac{1}{|O_\tau(s)|}  \left( \sum_{k\in O_\tau(t,s)} \frac{\varepsilon(t,s,k)}{\hat\varepsilon(s,k)} \right)
  \end{split}
\end{equation}
where
$O_\tau(s) = \{\ k\ |\ \exists t\ :\ X(t,s,k) \ne \emptyset\ \&\ v_k
\in V_\tau\ \}$ denotes the set of organisms that are contained in the
subtree $\tau$ and share at least one snoRNA of family $s$. $v_k$ is
the leaf that denotes organism $k$.


% --------------------------------------------------------------------------- %

\section{Results}
\input{results}
% --------------------------------------------------------------------------- %

\section{Discussion}
\input{discussion}

% --------------------------------------------------------------------------- %

\section*{Acknowledgments}

This work was funded in part by the European Union FP-7 project QUANTOMICS
(no.\ 222664), the MMML-seq project of the International Cancer Genome
Consortium (ICGC) funded by German Federal Ministry of Education and
Research, the CRC 1052 ``Obesity'', and by the Deutsche
Forschungsgemeinschaft (Project DFG STA 850/15-1). LIFE -- Leipzig Research
Center for Civilization Diseases is funded by the State of Saxony and the
European Union.

\bibliographystyle{m}
\bibliography{Fungi_snoRNAome}

\end{document}


